% Analysis 

```{r global_options, include = FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      comment = "##",
                      collapse = TRUE,
                      warning = FALSE,
                      message = FALSE)

```


# Libraries

To reproduce the examples of this material, the R packages the following packages are needed.

```{r warning=FALSE, message=FALSE}

library(EnvRtype)
library(rio)
library(ggh4x)
library(ggridges)
library(tidyverse)
library(metan)
library(lmerTest)
library(lme4)
library(broom)
library(broom.mixed)

my_theme <- 
  theme_bw() +
  theme(panel.spacing = unit(0, "cm"),
        panel.grid = element_blank(),
        legend.position = "bottom")
```


# Datasets
## Traits
```{r}
df_traits <- 
  import("https://bit.ly/df_barley") |> 
  as_factor(1:6)

```


## Climate variables
### Scripts to gather data

```{r eval=FALSE}
df_env <- import("https://bit.ly/3RvXlgj")
ENV <- df_env$env |> as_character()
LAT <- df_env$Lat
LON <- df_env$Lon
ALT <- df_env$Alt
START <- df_env$Sowing
END <- df_env$Harvesting

# see more at https://github.com/allogamous/EnvRtype
df_climate <-
  get_weather(env.id = ENV,
              lat = LAT,
              lon = LON,
              start.day = START,
              end.day = END)
# GDD: Growing Degree Day (oC/day)
# FRUE: Effect of temperature on radiation use efficiency (from 0 to 1)
# T2M_RANGE: Daily Temperature Range (oC day)
# SPV: Slope of saturation vapour pressure curve (kPa.Celsius)
# VPD: Vapour pressure deficit (kPa)
# ETP: Potential Evapotranspiration (mm.day)
# PEPT: Deficit by Precipitation (mm.day)
# n: Actual duration of sunshine (hour)
# N: Daylight hours (hour)
# RTA: Extraterrestrial radiation (MJ/m^2/day)
# SRAD: Solar radiation (MJ/m^2/day)
# T2M: Temperature at 2 Meters
# T2M_MAX: Maximum Temperature at 2 Meters
# T2M_MIN: Minimum Temperature at 2 Meters
# PRECTOT: Precipitation
# WS2M: Wind Speed at 2 Meters
# RH2M: Relative Humidity at 2 Meters
# T2MDEW: Dew/Frost Point at 2 Meters
# ALLSKY_SFC_LW_DWN: Downward Thermal Infrared (Longwave) Radiative Flux
# ALLSKY_SFC_SW_DWN: All Sky Insolation Incident on a Horizontal Surface
# ALLSKY_TOA_SW_DWN: Top-of-atmosphere Insolation
# [1] "env"       "ETP"       "GDD"         "PETP"      "RH2M"      "SPV"
# [8] "T2M"       "T2M_MAX"   "T2M_MIN"   "T2M_RANGE" "T2MDEW"    "VPD"

# Compute other parameters
env_data <-
  df_climate %>%
  as.data.frame() %>%
  param_temperature(Tbase1 = 5, # choose the base temperature here
                    Tbase2 = 28, # choose the base temperature here
                    merge = TRUE) %>%
  param_atmospheric(merge = TRUE) %>%
  param_radiation(merge = TRUE)
```

### Tidy climate data
```{r}
env_data <- import("https://bit.ly/env_data_nasapower")
str(env_data)
id_var <- names(env_data)[10:19]
```


# Scripts
## Correlation analysis of climate variables
```{r }
env_data |> 
  select_cols(all_of(id_var)) |> 
  corr_coef() |> 
  network_plot(min_cor = 0.3,
               legend_position = "bottom") 

ggsave("figs/fig1_network.png", width = 5, height = 5)

```


## Distribution of climate variables

```{r}
# long format for climate data
env_data_d <-
  env_data |>
  remove_cols(LON:YYYYMMDD, daysFromStart) |>
  pivot_longer(-env)

# long format for grain yield
df_gy_dist <- 
  df_traits |> 
  select(YEARS, GY) |> 
  mutate(name = "GY", .after = YEARS) |>
  rename(value = GY, 
         env = YEARS)
```


```{r}
# bind climate and GY
env_data_d <- rbind(df_gy_dist, env_data_d)
# mean values for each trait
env_data_mean <- means_by(env_data_d, name)

ggplot(env_data_d, aes(x = value, y = env, fill = env)) +
  geom_density_ridges(scale = 0.9) +
  geom_vline(data = env_data_mean,
             aes(xintercept = value),
             linetype = 2) +
  facet_grid(~name, scales = "free") +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank()) +
  scale_y_discrete(expand = expansion(c(0, 0.05))) +
  labs(x = "Observed value",
       y = "Density",
       fill = "") +
  my_theme

ggsave("figs/fig2_density_climate.png", width = 12, height = 5)
```



## Environmental tipology {.panelset}
```{r}
names.window <- c('1-initial growing',
                  '2-Tillering',
                  '3-Booting',
                  '4-Heading/Flowering',
                  "5-Grain filling",
                  "6-Physiological maturity")

out <- 
  env_typing(env.data = env_data,
             env.id = "env",
             var.id = c("trange", "tmax", "tmin", "dbp", "etp", "vpd"),
             by.interval = TRUE,
             time.window = c(0, 15, 35, 65, 90, 140),
             names.window = names.window)

out2 <- 
  separate(out,
           env.variable,
           into = c("var", "freq"),
           sep = "_",
           extra = "drop")

```


### tmax
```{r}
# plot the distribution of envirotypes for dbp
p1 <-
  out2 |> 
  subset(var == "tmax") |> # change the variable here
  ggplot() + 
  geom_bar(aes(x=Freq, y=interval,fill=freq), 
           position = "fill",
           stat = "identity",
           width = 1,
           color = "white",
           size=.2)+
  facet_grid(~env, scales = "free", space = "free")+ 
  scale_y_discrete(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0))+
  xlab('Relative frequency of the maximum temperature (ºC)')+ 
  ylab("Crop stage")+
  labs(fill='Envirotype')+
  theme(axis.title = element_text(size=12),
        legend.text = element_text(size=9),
        strip.text = element_text(size=12),
        legend.title = element_text(size=12),
        strip.background = element_rect(fill="gray95",size=1),
        legend.position = 'bottom')
```


### vpd
```{r}
# plot the distribution of envirotypes for dbp

p2 <-
  out2 |> 
  subset(var == "vpd") |> # change the variable here
  ggplot() + 
  geom_bar(aes(x=Freq, y=interval,fill=freq), 
           position = "fill",
           stat = "identity",
           width = 1,
           color = "white",
           size=.2)+
  facet_grid(~env, scales = "free", space = "free")+ 
  scale_y_discrete(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0))+
  xlab(expression(Relative~frequency~of~the~vapor~pressure~deficit~~(KPa~day^{-1})))+ 
  ylab("Crop stage")+
  labs(fill='Envirotype')+
  theme(axis.title = element_text(size=12),
        legend.text = element_text(size=9),
        strip.text = element_text(size=12),
        legend.title = element_text(size=12),
        strip.background = element_rect(fill="gray95",size=1),
        legend.position = 'bottom')

```



### dbp
```{r fig.cap="Quantiles for maximum temperature (a), vapor pressure deficit (b), and deficit by precipitation (c) across distinct crop stages of barley growing in two crop seasons.", fig.width=10, fig.height=8}
# plot the distribution of envirotypes for dbp

p3 <-
  out2 |> 
  subset(var == "dbp") |> # change the variable here
  ggplot() + 
  geom_bar(aes(x=Freq, y=interval,fill=freq), 
           position = "fill",
           stat = "identity",
           width = 1,
           color = "white",
           size=.2)+
  facet_grid(~env, scales = "free", space = "free")+ 
  scale_y_discrete(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0))+
  xlab(expression(Relative~frequency~of~the~rainfall~deficit~~(mm~day^{-1})))+ 
  ylab("Crop stage")+
  labs(fill='Envirotype')+
  theme(axis.title = element_text(size=12),
        legend.text = element_text(size=9),
        strip.text = element_text(size=12),
        legend.title = element_text(size=12),
        strip.background = element_rect(fill="gray95",size=1),
        legend.position = 'bottom')


arrange_ggplot(p1, p2, p3,
               ncol = 1,
               tag_levels = "a")
ggsave("figs/fig3_typology.png", width = 10, height = 9)
```


# Analysis 
## Model
```{r}
# Pivot longer to analyze each variable with group_by()
df_long <- 
  df_traits |> 
  pivot_longer(CAN:GY)

models <- 
  df_long %>% 
  group_by(YEARS, name) %>%
  doo(~lmer(value ~ STRESS + REP + (1|GENOTYPES) + (1|GENOTYPES:STRESS) + (1|REP:IBLOCK), data = .))
```



## LRT
```{r}
LRT <- 
  models |> 
  mutate(data = map(data,
                    ~.x %>% 
                      lmerTest::ranova(., reduce.terms = TRUE) |> 
                      rownames_to_column("term"))) |> 
  unnest(data)

```


## Variance components
```{r}

vcomp <- 
  models |> 
  mutate(data = map(data,
                    ~.x %>% 
                      lme4::VarCorr() |> 
                      as.data.frame() |>       
                      add_cols(Percent = (vcov / sum(vcov)) * 100))) |> 
  unnest(data)


ggplot(vcomp, aes(name, vcov, fill = grp)) +
  geom_bar(stat = "identity",
           position = "fill",
           color = "black",
           width = 1) +
  facet_wrap(~YEARS) +
  geom_text(aes(label = paste0(round(Percent, 2), "%")),
            position = position_fill(vjust = .5),
            size = 3) +
  scale_y_continuous(expand = expansion(c(0, 0)),
                     labels = function(x) paste0(x*100, "%")) +
  scale_x_discrete(expand = expansion(0)) +
  theme_bw()+
  theme(legend.position = "bottom",
        axis.ticks.length = unit(0.2, "cm"),
        panel.grid = element_blank(),
        legend.title = element_blank(),
        strip.background = element_rect(fill = NA),
        text = element_text(colour = "black"),
        axis.text = element_text(colour = "black")) +
  labs(x = "Assessed traits",
       y = "Proportion of phenotypic variance")

ggsave("figs/fig4_vcomp.png", width = 8, height = 6)
```


# BLUPs for Genotype x Stress by Year
```{r}

blups <- 
  models |> 
  mutate(data = map(data,
                    ~.x %>% 
                      augment())) |> 
  unnest(data) |> 
  group_by(YEARS, name, STRESS, GENOTYPES) |> 
  summarise(blup = mean(.fitted)) |> 
  ungroup() |> 
  pivot_wider(names_from = name, values_from = blup)

blups

```


## correlation between traits

```{r fig.cap="Correlation between traits under deficit irrigation (a) and full irrigation (b)"}
cor_di <- 
  blups |> 
  filter(STRESS == "DI") |> 
  corr_coef(CAN:PH) |> 
  network_plot(legend_position = "bottom", show = "all")

cor_fi <- 
  blups |> 
  filter(STRESS == "FI") |> 
  corr_coef(CAN:PH) |> 
  network_plot(legend_position = "bottom", show = "all")

arrange_ggplot(cor_di, cor_fi,
               guides = "collect",
               tag_levels = "a")

ggsave("figs/fig5_network_traits.png", width = 8, height = 5)
```




# MGIDI appied to each environment

```{r}
blups_ys <- split_factors(blups, YEARS, STRESS)

```

## 2019 DI
```{r}
mgidi_2019_di <- 
  blups_ys$`2019 | DI` |> 
  column_to_rownames("GENOTYPES") |> 
  mgidi(ideotype = c("l, h, h, h, l, h"),
        weights = c(1,1,1,5,1,2), # remove to consider equal heigths
        SI = 5,
        verbose = FALSE)
```

## 2019 FU
```{r}
mgidi_2019_fi <- 
  blups_ys$`2019 | FI` |> 
  column_to_rownames("GENOTYPES") |> 
  mgidi(ideotype = c("l, h, h, h, l, h"),
        weights = c(1,1,1,5,1,2), # remove to consider equal heigths
        SI = 5,
        verbose = FALSE)
```

## 2020 DI
```{r}
mgidi_2020_di <- 
  blups_ys$`2020 | DI` |> 
  column_to_rownames("GENOTYPES") |> 
  mgidi(ideotype = c("l, h, h, h, l, h"),
        weights = c(1,1,1,5,1,2), # remove to consider equal heigths
        SI = 5,
        verbose = FALSE)
```

## 2020 FU
```{r}
mgidi_2020_fi <- 
  blups_ys$`2020 | FI` |> 
  column_to_rownames("GENOTYPES") |> 
  mgidi(ideotype = c("l, h, h, h, l, h"),
        weights = c(1,1,1,5,1,2), # remove to consider equal heigths
        SI = 5,
        verbose = FALSE)
```



## Selected genotypes

```{r}
sel_2019DI <- mgidi_2019_di |> sel_gen()
sel_2019FI <- mgidi_2019_fi |> sel_gen()
sel_2020DI <- mgidi_2020_di |> sel_gen()
sel_2020FI <- mgidi_2020_fi |> sel_gen()

v1 <- 
  venn_plot(sel_2019DI, sel_2020DI,
            sel_2019FI, sel_2020FI,
            names = c("2019DI", "2020DI", "2019FI", "2020FI"))



# Selected in DI
DI <- set_union(sel_2019DI, sel_2020DI)
# Selected in FI
FI <- set_union(sel_2019FI, sel_2020FI)


v2 <- venn_plot(DI, FI)
arrange_ggplot(v1, v2,
               tag_levels = "a")

ggsave("figs/fig6_venn.png", width = 12, height = 5)

# only in DI (23)
set_difference(DI, FI)

# only in FI (18)
set_difference(FI, DI)


# in all environments
set_intersect(sel_2019DI, sel_2020DI, sel_2019FI, sel_2020FI)

```


## MGIDI index for the selected genotypes
```{r, fig.cap="Selected genotypes ranked by the MGIDI in the environments 2019DI (a), 2020DI (b), 2019FI (c), and 2020FI (d)"}

m1 <- 
  gmd(mgidi_2019_di, "MGIDI") |> 
  mutate(selected = ifelse(Genotype %in% set_union(sel_2019DI, sel_2020DI,
                                                   sel_2019FI, sel_2020FI),
                           "Yes",
                           "No")) |> 
  subset(selected == "Yes") |> 
  ggplot(aes(x = MGIDI, y = reorder(Genotype, -MGIDI))) +
  geom_line(aes(group = 1)) +
  geom_point(size = 2.5) +
  labs(x = "Multi-trait Genotype-Ideotype Distance Index",
       y = "Genotypes") +
  my_theme


m2 <- 
  gmd(mgidi_2019_fi, "MGIDI") |> 
  mutate(selected = ifelse(Genotype %in% set_union(sel_2019DI, sel_2020DI,
                                                   sel_2019FI, sel_2020FI),
                           "Yes",
                           "No")) |> 
  subset(selected == "Yes") |> 
  ggplot(aes(x = MGIDI, y = reorder(Genotype, -MGIDI))) +
  geom_line(aes(group = 1)) +
  geom_point(size = 2.5) +
  labs(x = "Multi-trait Genotype-Ideotype Distance Index",
       y = "Genotypes") +
  my_theme

m3 <- 
  gmd(mgidi_2020_di, "MGIDI") |> 
  mutate(selected = ifelse(Genotype %in% set_union(sel_2019DI, sel_2020DI,
                                                   sel_2019FI, sel_2020FI),
                           "Yes",
                           "No")) |> 
  subset(selected == "Yes") |> 
  ggplot(aes(x = MGIDI, y = reorder(Genotype, -MGIDI))) +
  geom_line(aes(group = 1)) +
  geom_point(size = 2.5) +
  labs(x = "Multi-trait Genotype-Ideotype Distance Index",
       y = "Genotypes") +
  my_theme

m4 <- 
  gmd(mgidi_2020_fi, "MGIDI") |> 
  mutate(selected = ifelse(Genotype %in% set_union(sel_2019DI, sel_2020DI,
                                                   sel_2019FI, sel_2020FI),
                           "Yes",
                           "No")) |> 
  subset(selected == "Yes") |> 
  ggplot(aes(x = MGIDI, y = reorder(Genotype, -MGIDI))) +
  geom_line(aes(group = 1)) +
  geom_point(size = 2.5) +
  labs(x = "Multi-trait Genotype-Ideotype Distance Index",
       y = "Genotypes") +
  my_theme

arrange_ggplot(m1, m2, m3, m4,
               ncol = 4,
               tag_levels = "a")

ggsave("figs/fig7_selected.png", width = 17, height = 8)

```



## Selection gains
```{r}
sel_gain <- 
  list(
    a = gmd(mgidi_2019_di) |> mutate(YEAR = 2019, STRESS = "Deficit irrigation"),
    b = gmd(mgidi_2019_fi) |> mutate(YEAR = 2019, STRESS = "Full irrigation"),
    c = gmd(mgidi_2020_di) |> mutate(YEAR = 2020, STRESS = "Deficit irrigation"),
    d = gmd(mgidi_2020_fi) |> mutate(YEAR = 2020, STRESS = "Full irrigation")
  ) |> 
  bind_rows() |> 
  mutate(negative = ifelse(SDperc <= 0, "Negative", "Positive"))

library(ggh4x)

ggplot(sel_gain, aes(VAR, SDperc)) +
  geom_hline(yintercept = 0, size = 0.2) +
  geom_col(aes(fill = negative),
           col = "black",
           size = 0.2) +
  scale_y_continuous(expand = expansion(mult = 0.1)) +
  facet_nested(~YEAR + STRESS +sense, scales = "free", space = "free") +
  geom_text(aes(label = round(SDperc, 2),
                hjust = ifelse(SDperc > 0, -0.1, 1.1),
                angle = 90),
            size = 3) +
  labs(x = "Traits",
       y = "Selection gains (%)") +
  my_theme +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(color = "black",
                                        fill = "white"))
ggsave("figs/fig8_selection_gains.png", width = 12, height = 6)
```



# Section info
```{r}
sessionInfo()
```

